Maximum Entropy Spectral Analysis: a case study

### 2021-06-07 paper/referee_response.md

- Please link to the correct citation at the start of Section VB, as [27] links to the climate data. "We focus on the public data released by the LIGO/Virgo collaboration [27]. We reconstruct the PSD both assuming the CAT and FPE loss functions on 1000s of data [27] from the Livingston observatory starting from GPS time 1164603392. The data are sampled at 4096 Hz.”
	+ Done

- "Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subtracting the expected signal with the actual signal.” Here the Bayeswave papers should be cited (1410.3835, 1410.3852) as well as other key LIGO papers describing glitches, their properties, and subtraction (1508.07316, 1602.03844, 1808.03619), and I believe [28] should point to the Zevin et. al paper in [29].
	+ Done: Bayeswave (as well as a study of its accuracy) is cited in sec.IV, page 10, together with 1907.06540 that assess its accuracy; papers on glitches are cited on sec. 5.2

- "for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA,” Please again add the citations for the LIGO and Virgo detector papers, and also add a citation to the KAGRA detector papers (1306.6747, 1111.7185).
	+ Done: pag 14

- "hence MESA could be employed for a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD”: Please add citations to other papers from LVC members discussing marginalization over PSD detector noise (0804.3853, 1109.0442, 1909.01934, 1409.7215, 1307.8195, 2004.05149, 2006.05292, 2101.01200, 2004.07515, 1506.00185 and the two Bayeswave papers mentioned above).
	+ Perhaps here there was a misundenstarding, as we did not make clear our idea. We do not aim to include in the PE the uncertainties of the PSD (as some works indicated by the referee do). Instead, our idea is to estimate the PSD at _every_ LL evaluation during a PE run. We tried to clarify this in the conclusion in the text and in a dedicated footnote.

- Please add a citation to https://arxiv.org/abs/gr-qc/0011041, which also demonstrates the applicability of autoregressive processes to gravitational-wave power spectral density estimation.
	+ Done: on sec. 5.2

- I’m not sure if this qualifies as a PnP comment, but a significant omission is the lack of discussion of the Bayeswave method of PSD estimation, which is the official method for calculating PSDs for parameter estimation results by the LVC. In addition to citing the papers described above, I think it would be appropriate to add a discussion, either in the Introduction or the Final Remarks comparing MESA to this “on-source” PSD estimation and citing the relvant papers disucssing the Bayeswave PSD method in comparison to the Welch method, some repeated from the above requests (1907.06540, 2006.05292).
	+ We found that the comparison with Bayesline goes Beyond the scope of our work. Bayesline is a parametric model (unlike the non parametric Welch and MESA) and it is taylored to GW data analysis. Our discussion here aimed to be inspired from but broader than GW data analysis. Thus, a comparison with BayesLine would narrow too much the applicability and the interest of our work.
	We mentioned this point in a newly added paragraph in the introduction (where we cited the two papers suggested).

@@

## Personal comments

We fixed all the typos that the referee pointed out. We are thankful for that.

### Validations of the model:
- I wasn't able to get a good grasp of the practical applications of the estimation of the autoregressive process order while reading this section. Is gravitational-wave data expected to be stationary enough to reliably model as an autoregressive process? If not, what are some other examples? It would be very helpful to elaborate on this, perhaps when the connection to AR is introduced.
	+ We added at the beginning of the relevant section (3.4) a note on this, trying to explain the importance of such study.

@@

### Comparison with Welch method:
- Maybe it’s because of the broken citation, but I don’t understand which PSD is being used in the simulation. Is it the same “analytical PSD computed for the LIGO Hanford interferometer” used in the previous section, or is it a PSD calculated using the real LIGO data from around GW150914? If so, how was that PSD calculated?
	+ The two PSDs are different. In the validation part, we used the design sensitivity of aLIGO (we quoted the relevant paper as well as the DCC link for the file). In the Welch comparison part, we employed the PSD released together with GW150914 (again we quoted the GWTC-1 paper and the link).

### Applications:

- It’s not really clear to me when you perform the forecasting for the temperature data why CAT is the obvious choice for the loss function, since in the previous section you described how CAT is unable to recover the true order of the autoregressive process. Does that mean that it would be worse at forecasting based on the AR(p)? There seems to be a contradiction where CAT does a better job for the temperature data but worse for the simulation presented in section II. It would be helpful to elaborate on this.
	+ We added a claryfing footnote on this (note 10, page 10). We hope it is clearer now.

- "Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal.”: How does this compare to the method presented in https://arxiv.org/abs/1908.05644, where the data containing a glitch are “inpainted” with values chosen such that the inverse-PSD-filtered data is 0 at these times?
	+ To answer this matter is not easy and this should be addressed in future works. Thanks for pointing out the reference: we added it

### 2021-06-07 paper/mesa.tex
Some edits to the paper

An appealing alternative approach, based on the Maximum Entropy principle \citep{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been put forward by~\citet{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

@@

We now address the issue of how well the AR order (i.e. the number of $a_k$ coefficients employed) is estimated by each loss function.
In doing so, we generate $100$ autoregressive processes $AR(p)$ with a random value of $p$, drawn such that $log(p) \sim\mathcal{U}_{[log(2), log(5000)]}$.
Each coefficient $a_k$ is assigned according to a Dirichlet distribution $\mathrm{Dir}([1,\hdots, 1])$. The sign of $a_k$ is assigned randomly according to a binomial distribution. We report the result of this investigation in Fig.~\ref{fig:p_vs_ptrue}.

@@

\sschmidt{How does this compare to the method presented in https://arxiv.org/abs/1908.05644? Seems like an interesting paper that tells you how to produce data after glitch removal.}
Despite poor predictions on long timescales, the AR process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal. Indeed loud anomalies in the data, called {\it glitches}, pose a major challenge to the ability of detecting signals and intensive work has been done to mitigate the disruption in sensitivity they cause \citep{Nuttall_2015, Noise_char_2016, Zevin_2017} and to develop effective subraction techniques \citep{Pankow_2018}.
The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subtracting the expected signal with the actual signal.
This can work: a typical glitch can last as long as $\SI{0.2}{s}$, close to the time scale over which the forecasting is reliable.

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature \citep[see]{LSTM_article, oord2016wavenet} and they can be trained for precision forecasting. However, a Maximum-Entropy trained AR process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA \cite{LIGO_2015, Virgo_2014, KAGRA_2012, KAGRA_2013}, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for online PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \cite{lalinference}: this scheme assumes {\it stationary data} on long timescale and it might not reflect the structure of the noise in the time slice to analyse.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \cite{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \cite{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making the stationariety assumption as minimal as possible.
}, getting, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\cite{martini_thesis}.

>>>>

\sschmidt{This could be a footnote?}\textbf{A different approach, which provides a smooth PSD estimation, is to set up a parametric model for the PSD and to fit its parameters to the data with a Markov chain Monte Carlo \cite{Cornish_2015, Littenberg_2015}. Despite being effective, such method is very problem dependent, as it makes heavy assumption on the shape of the PSD. Moreover, it is computationally expensive and it does not come with an handy implementation available to the public. For all those reasons, we do not consider such method in our work.}

An appealing alternative, based on the Maximum Entropy principle \citep{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been put forward by~\citet{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

@@

We now address the issue of how well the AR order (i.e. the number of $a_k$ coefficients employed) is estimated by each loss function.
This is useful to further characterize the properties of the different loss functions. Furthermore, as MESA is supposed to model data as an $AR(p)$ process, it is interesting to quantify its precision on some true autoregressive (and stationary) time series.

We generate $100$ autoregressive processes $AR(p)$ with a random value of $p$, drawn such that $log(p) \sim\mathcal{U}_{[log(2), log(5000)]}$.
Each coefficient $a_k$ is assigned according to a Dirichlet distribution $\mathrm{Dir}([1,\hdots, 1])$. The sign of $a_k$ is assigned randomly according to a binomial distribution. We report the result of this investigation in Fig.~\ref{fig:p_vs_ptrue}.

@@

Despite poor predictions on long timescales, the AR process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal. Indeed loud anomalies in the data, called {\it glitches}, pose a major challenge to the ability of detecting signals and intensive work has been done to mitigate the disruption in sensitivity they cause \citep{Nuttall_2015, Noise_char_2016, Zevin_2017} and to develop effective subraction techniques \citep{Pankow_2018, zackay2019detecting}.
The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subtracting the expected signal with the actual signal.
This can work: a typical glitch can last as long as $\SI{0.2}{s}$, close to the time scale over which the forecasting is reliable. Future works will explore this exicting opportunity.

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature (see e.g. \cite{LSTM_article, oord2016wavenet}) and they can be trained for precision forecasting. However, a Maximum-Entropy trained AR process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \cite{lalinference}: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \cite{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \cite{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\cite{martini_thesis}.
